Viscous Erosion in Porous Media

### 2018-03-03 docs/methods/erosion_methods.tex
NM
Put final touches on theta-L and timestepping sections.

The boundary of each body will be represented by a grid of collocation points. If we were to simply evolve the interfaces under \eqref{eqn:Vn2}, this grid would become distorted, which, if unchecked, could lead to severe tangling and numerical instability. We therefore introduce an artificial tangential velocity $\Vs$ that always keeps the grid points equally spaced in arclength \cite{hou-low-she1994}. The appropriate tangential velocity satisfies 
\begin{align}
\pderiv{\Vs}{\alpha} = \thalpha \Vn - \mean{\thalpha \Vn} \, ,
\end{align}
The evolution of the interface is then given by the combination of the normal and tangential velocity
\begin{align}
\dot{\xx}(t) = \Vn \nn + \Vs \ss \, .
\end{align}
Since $\Vs$ is tangential, it does not modify the erosion law \eqref{eqn:Vn}, only the collocation grid.

When recast in the variables $\theta$ and $L$, the interface evolution becomes
\begin{align}
\tderiv{L}{t} &= - 2 \pi \mean{\thalpha \Vn}, \\
\pderiv{\theta}{t} &= \frac{2 \pi}{L} \left( \pderiv{\Vn}{\alpha} +
\thalpha \Vs \right).
\end{align}
Inserting~\eqref{eqn:Vn2} for the normal velocity gives
\begin{align}
\label{eqn:Levo1}
\tderiv{L}{t} &= - 2 \pi \mean{\thalpha \Vn}, \\
\label{eqn:thetaevo1}
\pderiv{\theta}{t} &= \frac{2\pi}{L} \left(
\eps \mean{\atau} \ppd{\theta}{\alpha} + \pderiv{\atau}{\alpha} +
\thalpha \Vs \right).
\end{align}
System \eqref{eqn:Levo1}--\eqref{eqn:thetaevo1} comprise the main evolution equations to be solved numerically in Section~\ref{sec:timeStepping}.

In addition, it is necessary to track a reference point for each body, so that the variables $\theta$ and $L$ can be converted back to cartesian coordinates. We track the coordinates of the surface mean $\mean{\xx}$, which moves according to
\begin{align}
\label{eq:xsurf}
\frac{d}{dt} \mean{\xx} = \mean{ V_s \ss + V_n \nn }
\end{align}
We note that tracking the center of mass is also a valid choice, but the surface mean gives a somewhat simpler formula.

@@

With the shear stress computed, we are now ready to evolve the boundary of each body using equations~\eqref{eqn:Levo1} and~\eqref{eqn:thetaevo1}.  As described in Section~\ref{sec:thetaL}, we use the {\thL} framework, with $\theta$ being the tangent angle and $L$ the total arc length of a given boundary. Each boundary is parameterized by normalized arc length $\alpha = 2 \pi s / L$.

\todo[inline]{Mention key term exponential integrator for stiff term.}

In addition to the curvature-dependent term in~\eqref{eqn:Vn}, we regularize boundary evolution by applying a narrow Gaussian filter to absolute shear stress
\begin{align*}
\atausig = \int_{\gamma} \frac{1}{\sqrt{2\pi} \sigma} \exp \left(-
\frac{\alpha'^2}{2 \sigma^2}\right) \abs{ \tau(\alpha - \alpha') } \,
d\alpha'.
\end{align*}
Here $\sigma$ sets the width of the filter. The corresponding normal and tangential interface velocities are then given by
\begin{align*}
& \Vnsig = \atausig +  \eps \mean{\atausig}
\left(\pderiv{\theta}{\alpha} - 1 \right), \\
& \pderiv{\Vssig}{\alpha} = \thalpha \Vnsig - \mean{\thalpha \Vnsig}.
\end{align*}

For convenience, we introduce the variables
\begin{align*}
%\label{eq:M}
& \Mterm = - 2 \pi \mean{\thalpha \Vnsig}, \\
%\label{eq:N}
& \NLterm = \frac{2 \pi}{L} \left( \pderiv{\atausig}{\alpha} + \thalpha
\Vssig \right), \\
%\label{eq:elfun}
& \elfun = \frac{2 \pi}{L}  \mean{\atausig} \, .
\end{align*}
Then, the evolution equations~\eqref{eqn:Levo1}
and~\eqref{eqn:thetaevo1} can be compactly expressed as
\begin{align}
\label{eqn:Levo2}
& \tderiv{L}{t} = \Mterm, \\
\label{eqn:thetaevo2}
& \pderiv{\theta}{t} = \eps \elfun \ppd{\theta}{\alpha} + \NLterm.
\end{align}
Here $\NLterm$ represents the nonlinear terms (and nonlocal terms) in the PDE for $\theta$.

\todo[inline]{Need to remind reader that the diffusive term comes from smoothing and that this is the stiff term that could potential cause numerical problems.}

We will evolve $\theta(\alpha)$ spectrally and so introduce the Fourier series
\begin{align}
\label{thetaSeries}
 \theta(\alpha,t) &= \alpha + \FourierSum \thhat_k e^{ i k \alpha}, \\
 \NLterm(\alpha,t)  &= \FourierSum \widehat{\NLterm}_k e^{ i k \alpha}.
\end{align}
The careful reader will notice that in \eqref{thetaSeries}, we have decomposed $\theta(\alpha)$ into a linear component and a periodic component.
In spectral space,~\eqref{eqn:thetaevo2} produces the system of ODEs
\begin{align*}
\tderiv{\thhat_k}{t} +  \eps k^2  \elfun \thhat_k = \widehat{\NLterm}_k.
\end{align*}
These ODEs can be solved by using the integrating factor
\begin{align}
\label{eq:mu}
\mu_k = \exp \left( \eps k^2 \int \elfun \, dt \right).
\end{align}
We are therefore led to the coupled system of ODEs
\begin{align}
\label{eqn:Levo3}
& \tderiv{L}{t} = \Mterm, \\
\label{eqn:thetaevo3}
& \tderiv{}{t}\left( \mu_k \thhat_k \right) = \mu_k \widehat{\NLterm}_k \, ,
\qquad k = -N_\iin/2, \ldots, N_\iin/2 -1
\end{align}
We solve system~\eqref{eq:mu}--\eqref{eqn:thetaevo3} with an
explicit, second-order Runge-Kutta method (RK2). In particular, we choose the {\em midpoint} RK2 method so as to maximize the smoothing effect of the curvature term.

>>>>

The boundary of each body will be represented by a grid of collocation points. If we were to simply evolve the interfaces under \eqref{eqn:Vn2}, this grid would become distorted, which, if unchecked, could lead to severe tangling and numerical instability. We therefore introduce an artificial tangential velocity $\Vs$ that always keeps the grid points equally spaced in arclength \cite{hou-low-she1994}. The appropriate tangential velocity satisfies 
\begin{align}
\pderiv{\Vs}{\alpha} = \thalpha \Vn - \mean{\thalpha \Vn} \, ,
\end{align}
The evolution of the interface is then given by the combination of the normal and tangential velocity
\begin{align}
\dot{\xx}(t) = \Vn \nn + \Vs \ss \, .
\end{align}
Since $\Vs$ is tangential, it does not modify the physics of erosion, only the grid spacing.

When recast in the variables $\theta$ and $L$, the interface evolution becomes
\begin{align}
\tderiv{L}{t} &= - 2 \pi \mean{\thalpha \Vn}, \\
\pderiv{\theta}{t} &= \frac{2 \pi}{L} \left( \pderiv{\Vn}{\alpha} +
\thalpha \Vs \right).
\end{align}
Inserting~\eqref{eqn:Vn2} for the normal velocity gives
\begin{align}
\label{eqn:Levo1}
\tderiv{L}{t} &= - 2 \pi \mean{\thalpha \Vn}, \\
\label{eqn:thetaevo1}
\pderiv{\theta}{t} &= \frac{2\pi}{L} \left(
\eps \mean{\atau} \ppd{\theta}{\alpha} + \pderiv{\atau}{\alpha} +
\thalpha \Vs \right).
\end{align}
These are the main evolution equations for $\theta(\alpha,t)$ and $L(t)$.
In addition, it is necessary to track a reference point for each body, so that the {\thL} variables can be converted back to cartesian coordinates. We track the coordinates of the surface mean $\mean{\xx}$, which moves according to
\begin{align}
\label{eq:xsurf}
\frac{d}{dt} \mean{\xx} = \mean{ V_s \ss + V_n \nn }
\end{align}
We note that tracking the center of mass is also a valid choice, but the surface mean gives a somewhat simpler formula. Equations \eqref{eqn:Levo1}--\eqref{eq:xsurf} comprise the main system of equations to be solve numerically in Section~\ref{sec:timeStepping}.

@@

With the shear stress computed, we are now ready to numerically evolve the boundaries using the {\thL} framework. The main equations to be solved are \eqref{eqn:Levo1}--\eqref{eq:xsurf}.  In Section~\ref{sec:thetaL} we introduced a curvature-penalization term into the erosion law \eqref{eqn:Vn}. This term smooths out high-frequency oscillations, but also introduces {\em stiffness} into evolution equation \eqref{eqn:thetaevo1} for $\theta(\alpha,t)$. For numerical stability, we will treat the stiff term with an exponential integrator, while handling all other terms in \eqref{eqn:Levo1}--\eqref{eq:xsurf} with an explicit Runge-Kutta method. Ultimately, we will arrive at a method that is second-order in time.

First, we must address a second potential source of numerical instability. The erosion law \eqref{eqn:Vn} depends on the {\em absolute value} of shear stress, $\atau$, which exhibits a corner anywhere $\tau$ changes sign (a stagnation point). This corner in $\atau$ introduces high frequencies into \eqref{eqn:thetaevo1} which could result in either numerical instability or an extremely stringent step-size restriction. To avoid these difficulties, we smooth $\atau$ via a narrow Gaussian filter of width $\sigma$
\begin{align}
\atausig = \int_{\gamma} \frac{1}{\sqrt{2\pi} \sigma}
 \exp \left(- \frac{\alpha'^2}{2 \sigma^2}\right) \abs{ \tau(\alpha - \alpha') } \, d\alpha' \, ,
\end{align}
with corresponding normal and tangential interface velocities
\begin{align}
& \Vnsig = \atausig +  \eps \mean{\atausig}
\left(\pderiv{\theta}{\alpha} - 1 \right), \\
& \pderiv{\Vssig}{\alpha} = \thalpha \Vnsig - \mean{\thalpha \Vnsig}.
\end{align}

We now consider the system of evolution equations given by \eqref{eqn:Levo1}--\eqref{eq:xsurf}. For convenience, we introduce the following variables
\begin{align}
%\label{eq:M}
& \Mterm = - 2 \pi \mean{\thalpha \Vnsig}, \\
%\label{eq:N}
& \NLterm = \frac{2 \pi}{L} \left( \pderiv{\atausig}{\alpha} + \thalpha
\Vssig \right), \\
%\label{eq:elfun}
& \elfun = \frac{2 \pi}{L}  \mean{\atausig} \, .
\end{align}
Then, ~\eqref{eqn:Levo1} and~\eqref{eqn:thetaevo1} can be compactly expressed as
\begin{align}
\label{eqn:Levo2}
& \tderiv{L}{t} = \Mterm, \\
\label{eqn:thetaevo2}
& \pderiv{\theta}{t} = \eps \elfun \ppd{\theta}{\alpha} + \NLterm.
\end{align}
The first term on the right-hand-side of \eqref{eqn:thetaevo2} is the stiff term that results from the curvature penalization. This term is multiplied by $\zeta$, which is spatially constant but time dependent (due to $L$ and $\mean{\atau}$). The second term, $\NLterm$, represents the rest of the evolution equation, including contributions from the shear-dependent erosion law and from the artificial tangential velocity. Since the local shear stress depends on interactions between bodies, $\NLterm$ is both a nonlinear and nonlocal.

We will evolve $\theta(\alpha,t)$ spectrally and so introduce the Fourier series
\begin{align}
\label{thetaSeries}
 \theta(\alpha,t) &= \alpha + \FourierSum \thhat_k e^{ i k \alpha} \, , \\
 \NLterm(\alpha,t)  &= \FourierSum \widehat{\NLterm}_k e^{ i k \alpha} \, .
\end{align}
The careful reader will notice that in \eqref{thetaSeries}, we have decomposed $\theta(\alpha)$ into a linear component and a periodic component. In spectral space,~\eqref{eqn:thetaevo2} produces the system of ODEs
\begin{align*}
\tderiv{\thhat_k}{t} +  \eps k^2  \elfun \thhat_k = \widehat{\NLterm}_k \, .
\end{align*}
These ODEs can be solved by applying the integrating factor
\begin{align}
\label{eq:mu}
\mu_k = \exp \left( \eps k^2 \int \elfun \, dt \right) \, ,
\end{align}
which leads to the coupled system of ODEs
\begin{align}
\label{eqn:Levo3}
& \tderiv{L}{t} = \Mterm, \\
\label{eqn:thetaevo3}
& \tderiv{}{t}\left( \mu_k \thhat_k \right) = \mu_k \widehat{\NLterm}_k \, ,
\qquad k = -N_\iin/2, \ldots, N_\iin/2 -1
\end{align}
We solve system~\eqref{eq:mu}--\eqref{eqn:thetaevo3} with an
explicit, second-order Runge-Kutta method (RK2). In particular, we choose the {\em midpoint} RK2 method so as to maximize the smoothing effect of the curvature term.
